Variants of G\"{o}del's Ontological Proof \\ in a Natural Deduction Calculus

### 2016-03-30 Papers/InProgress/NaturalDeduction/GodProof-ND.tex
Change \psi to \phi
Improve text

Without the restriction that every box must be accessed by exactly one strong modal inference, the calculus would be unsound for the modal logic \textbf{K}. For example, the formula $\all \psi. (\nec \psi \imp \pos \psi)$ is not valid in \textbf{K} but would be provable without this restriction:

\begin{prooftree}
  \AXC{$ $} \RightLabel{1}
  \UIC{$\nec \psi$} \RightLabel{$\nec_E$}
        \UIC{$\omega: \psi$} \RightLabel{$\pos_I$}
        \UIC{$\pos \psi$} \RightLabel{$\imp_I^1$}
    \UIC{$\nec \psi \imp \pos \psi$} \RightLabel{$\forall_I$}
    \UIC{$\all \psi. (\nec \psi \imp \pos \psi)$}
\end{prooftree}

@@

The straightforward combinations of the quantifier rules of {\ND} either with the modal rules of {\NDK} or with axiom K and the necessitation rule are suitable for a higher-order modal logic where constants and variables are \emph{rigid}. From the point of view of a \emph{possible world semantics}, rigidity means that their interpretation is independent of the world in which they are being interpreted. Rigidity is silently assumed by most works investigating the ontological argument, and is explicitly assumed here. Nevertheless, it should be noted that its adequacy for the ontological argument has already been contested \citep{fitting}. Another assumption made here is that the quantification domains are constant (i.e. independent of the possible worlds). Neither G\"odel's manuscript nor Scott's manuscript reveal whether they use constant or varying domains, and this is also the case for many variants (e.g. \citep{hajek}). Nevertheless, some authors of variants \citep[footnotes 11 and 14]{and90} of G\"odel's ontological argument do explicitly state a preference for varying domains. Our choice of assuming constant domains is motivated by simplicity.

@@

From an axiomatic point of view, modal logics differ with respect to which additional axiom schemas they admit. Some common axiom schemas of relevance to the ontological argument are T ($\nec A \imp A$), B ($A \imp \nec \pos A$), 4 ($\nec A \imp \nec \nec A $) and 5 ($\pos A \imp \nec \pos A$). From a semantic point of view, these axioms correspond to geometric frame conditions that must be satisfied by the accesibility relation in the Kripke models \citep{negri}: T corresponds to reflexivity, B corresponds to symmetry, 4 corresponds to transitivity and 5 corresponds to euclidianity. In the models of logic {\bf S5}, all these axioms are satisfied; in the models of logic {\bf KB}, K and B are satisfied; and in the models of logic {\bf B}, K, T and B are satisfied.

@@

\noindent
\emph{Brouwer's reduction principle} is derivable in modal logic {\bf KB}, using axiom B. The proof of G\"odel and Scott (Section \ref{sec:scott}) do not make direct use of this principle, but the new proof presented in Section \ref{sec:newproof} does.

@@

Scott's version of G\"odel's proof depends on 5 axioms that circumscribe the notion of \emph{positive} property, with positivity being denoted by the undefined second-order predicate symbol $P$. Additionally, 3 definitions are used for defining the notions of \emph{God-like}, \emph{essence} and \emph{necessary existence}. Technically, these definitions only abbreviate certain complex formulas. The argument would still go through if all defined symbols were replaced by the complex formulas they define. This observation is particularly relevant in the case of \emph{necessary existence}. Since this notion of ``existence'' is just an abbreviation, G\"odel's argument is not susceptible to Kant's criticism against Anselm's argument (that existence should not be treated as a predicate). In G\"odel's proof, existence is properly denoted by the existential quantifier. It is, therefore, unfortunately misleading to refer to the defined predicate symbol $E$ as ``necessary existence'', when in fact it is just a convenient abbreviation.

Beside the 5 axioms and 3 definitions that pertain specifically to the argument, the proof also uses {\bf S5}'s iteration principle (in the proof of Lemma 2) and the equality axiom of reflexivity (in the proof of Theorem 1). The use of reflexivity could be avoided, but allows a shorter proof.

@@

\section{New Proof}\label{sec:newproof}

In this section we present a new proof in {\bf KB}. This proof uses exactly the same axioms and definitions of Scott's proof shown in the previous sections, but it uses neither the equality axiom of reflexivity nor {\bf S5}'s modal iteration principle. Instead, it relies only on Brouwer's reduction principle (in the proof of Theorem 3). The new proof is also shorter than Scott's proof.

@@

Although the simpler proof of Theorem 1 presented here was developed independently, it seems conceptually related to a possible simplification briefly and informally described by Anderson \citep[footnote 2]{and90} and attributed by him to an anonymous referee. The simpler proof of Theorem 1 is not particularly special. In fact, it is probably the most likely proof that any well-trained logician would construct, if asked to derive Theorem 1 from Axioms 1 and 2 using methodical formal proof search techniques. Modern techniques usually rely on unification to compute or guess good instances for weakly quantified variables, and therefore they would never result in instantiations containing $=$ and $\neq$, which are symbols that do not appear anywhere in the axioms or in the theorems to be proved. We may only wonder why G\"odel preferred to use those unusual instantiations. It might be an interesting question from a historical point of view, if this suggests that there might have been a change (perhaps evolution) in the way logicians construct formal proofs; or even from a philosophical point of view, if G\"odel's preference was philosophically motivated.

>>>>

Without the restriction that every box must be accessed by exactly one strong modal inference, the calculus would be unsound for the modal logic \textbf{K}. For example, the formula $\all \phi. (\nec \phi \imp \pos \phi)$ is not valid in \textbf{K} but would be provable without this restriction:

\begin{prooftree}
  \AXC{$ $} \RightLabel{1}
  \UIC{$\nec \phi$} \RightLabel{$\nec_E$}
        \UIC{$\omega: \phi$} \RightLabel{$\pos_I$}
        \UIC{$\pos \phi$} \RightLabel{$\imp_I^1$}
    \UIC{$\nec \phi \imp \pos \phi$} \RightLabel{$\forall_I$}
    \UIC{$\all \phi. (\nec \phi \imp \pos \phi)$}
\end{prooftree}

@@

The straightforward combinations of the quantifier rules of {\ND} either with the modal rules of {\NDK} or with axiom K and the necessitation rule are suitable for a higher-order modal logic where constants and variables are \emph{rigid}. From the point of view of a \emph{possible world semantics}, rigidity means that their interpretation is independent of the world in which they are being interpreted. Rigidity is tacitly assumed by most works investigating the ontological argument, and is explicitly assumed here. Nevertheless, it should be noted that its adequacy for the ontological argument has already been contested \citep{fitting}. Another assumption made here is that the quantification domains are constant (i.e. independent of the possible worlds). Neither G\"odel's manuscript nor Scott's manuscript reveal whether they use constant or varying domains, and this is also the case for many variants (e.g. \citep{hajek}). Nevertheless, some authors of variants \citep[footnotes 11 and 14]{and90} of G\"odel's ontological argument do explicitly state a preference for varying domains. Our choice of assuming constant domains is motivated by simplicity.

@@

From an axiomatic point of view, modal logics differ with respect to which additional axiom schemas they admit. Some common axiom schemas of relevance to the ontological argument are T ($\nec A \imp A$), B ($A \imp \nec \pos A$), 4 ($\nec A \imp \nec \nec A $) and 5 ($\pos A \imp \nec \pos A$). From a semantic point of view, these axioms correspond to geometric frame conditions that must be satisfied by the accessibility relation in Kripke models \citep{negri}: T corresponds to reflexivity, B corresponds to symmetry, 4 corresponds to transitivity and 5 corresponds to euclidianity. In the models of logic {\bf S5}, all these axioms are satisfied; in the models of logic {\bf KB}, K and B are satisfied; and in the models of logic {\bf B}, K, T and B are satisfied.

@@

\noindent
\emph{Brouwer's reduction principle} is derivable in modal logic {\bf KB}, using axiom B. The proofs of G\"odel and Scott (Section \ref{sec:scott}) do not make direct use of this principle, but the new proof presented in Section \ref{sec:newproof} does.

@@

Scott's version of G\"odel's proof depends on five axioms that circumscribe the notion of \emph{positive} property, with positivity being denoted by the undefined second-order predicate symbol $P$. Additionally, three definitions are used for establishing the notions of \emph{God-like}, \emph{essence} and \emph{necessary existence}. Technically, these definitions only abbreviate certain complex formulas. The argument would still go through if all defined symbols were replaced by the complex formulas they define. This observation is particularly relevant in the case of \emph{necessary existence}. Since this notion of ``existence'' is just an abbreviation, G\"odel's argument is not susceptible to Kant's criticism against Anselm's argument (that existence should not be treated as a predicate). In G\"odel's proof, existence is properly denoted by the existential quantifier. It is, therefore, unfortunately misleading to refer to the defined predicate symbol $E$ as ``necessary existence'', when in fact it is just a convenient abbreviation.

In addition to the five axioms and three definitions that pertain specifically to the argument, the proof also uses {\bf S5}'s iteration principle (in the proof of Lemma 2) and the equality axiom of reflexivity (in the proof of Theorem 1). The use of reflexivity could be avoided, but we chose to use it because it allows the construction of a shorter proof.

@@

\section{A New Proof}\label{sec:newproof}

In this section we present a new proof in {\bf KB}. This proof uses exactly the same axioms and definitions of Scott's proof shown in the previous sections, but it uses neither the equality axiom of reflexivity nor {\bf S5}'s modal iteration principle. Instead, it relies only on Brouwer's reduction principle (for Theorem 3). The new proof is also shorter than Scott's proof.

@@

Although the simpler proof of Theorem 1 presented here was developed independently, it seems conceptually related to a possible simplification briefly described by Anderson \citep[footnote 2]{and90} and attributed by him to an anonymous referee. The simpler proof of Theorem 1 is not particularly special. In fact, it is probably the most likely proof that any well-trained logician would construct, if asked to derive Theorem 1 from Axioms 1 and 2 using methodical formal proof search techniques. Modern techniques usually rely on unification to compute or guess good instances for weakly quantified variables, and therefore they would never result in instantiations containing $=$ and $\neq$, which are symbols that do not appear anywhere in the axioms or in the theorems to be proved. We may wonder why G\"odel preferred to use those unusual instantiations. It might be an interesting question from a historical point of view, as it might suggest that there was an evolution in the way logicians construct formal proofs; or from a philosophical point of view, if G\"odel's preference was philosophically motivated.
